

# Load libraries and set up chunks options.
```{r setup, include=F, warnings=F, echo=F, message=F}
library(knitr)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(caret)
library(kernlab)
library(class)
library(randomForest)
library(keras)
library(C50) 

knitr::opts_chunk$set(echo = T, warning = F, include=T, message = F,
                      results = 'show')
knitr::opts_knit$set(root.dir = normalizePath("~/ML"))
```

# Data loading and preparation
## Read and explore the data
Data is already normalized.
```{r load_data}
rna <- read.csv("rna.csv")

# Convert the cancer types to factor
rna$Y <- factor(rna$Y)
prop.table(table(rna$Y))
```
In this case, the distribution of cancer types is not equitative.


```{r stat_descriptive}
# We create a dataframe to store different parameters of the data
set.seed(22)
df <- data.frame(
      Parámetro=c("Cancer type", "Genes"),
      Number=c(length(unique(rna$Y)), length(rna[,-1])),
      Content=c(paste(unique(rna$Y), collapse=", "),
                  # para los genes indicamos solo algunos
                  paste(c(names(rna)[sample(2:length(rna),5,)], "..."),
                        collapse=", "))
      )
df
```

We also explore data with a heatmap

```{r heatmap, fig.width=18, fig.height=12}
# heatmap
ph <- rna[,-1] %>% t()
colnames(ph) <- rna$Y
pheatmap(ph ,
         scale = "row",
         rev(brewer.pal(8, ("RdBu"))),
         show_rownames = F,
         cluster_rows = T, 
         cluster_cols = T
        )
```

## Split dataset into train and test
We apply the same splitting of traning and test data to be used in all the algorithms. We set the traning dataset with 75% of the samples and the rest for train (25%).


```{r split_set}
set.seed(22)

# We use Caret function to ensure the same factor distribution in test and train
part <- createDataPartition(rna$Y, p = 0.75,
                              list = FALSE)
train0 <- rna[part, ]
test0 <- rna[-part, ]

prop.table(table(train0$Y))
prop.table(table(test0$Y))
```



# Algorithms application
We create a list to store all the models and a dataframe for their evaluation, in order to compare them all.
```{r}
models <- list() 

df_cm <- data.frame(matrix(nrow=0,ncol=4))
names(df_cm) <- c("metric", "value", "class", "Model")
```
We create a function to extract the evaluation results with the confusuionMatrix() function from Caret.
```{r cm_results}
store_cm <- function(cm, model){
  overall <- data.frame(metric = names(cm$overall[1:2]),
                        value = cm$overall[1:2] %>% as.numeric(),
                        class = "Overall",
                        Model = model)
  byclass <- cm$byClass %>% as.data.frame() %>%
    rownames_to_column(var="class") %>% 
    pivot_longer(!class, names_to="metric", values_to="value")
  byclass$Model <- model
  
  df0 <- rbind(overall, byclass)
  return(df0)
}
```

Also, let's create a function to show tables with kappa and accuracy values.

```{r acc_kappa}
sum_AcK <- function(model, df_cm=df_cm){
  df <- df_cm %>% filter(grepl(model, Model, ignore.case = T),
                 class == "Overall") %>% 
          select(-class) %>% 
          pivot_wider(names_from = Model, values_from = value) %>% 
          as.data.frame()
  pander::pander(df, style="grid", justify="left")
}
```



## k-Nearest Neighbour

Let's explore k-Nearest Neighbour for values k = 1, 3, 5, 7, 11.

### Data transformation
Data is alrady normalized, we only need to scale it.

```{r}
train <- train0
test <- test0
train[,-1] <- scale(train[,-1])
test[,-1] <- scale(test[,-1])
```

### Model training
We will use the function `kNN` from `Caret` package.


```{r k-nearest}
ks <- c(1,3,5,7,11) # k values to test

set.seed(22)
tr_control <- trainControl(method="repeatedcv", # method crossvalidation
                           repeats=3) # 3-fold crossvalidation

knn <- train(Y ~ ., data = train,
             method = "knn",
             trControl = tr_control,
             tuneGrid = data.frame(k=ks))
knn
models[["kNN"]] <- knn

# plot
plot(knn)
```


### Prediction and evaluation of the model
```{r knn_eval}
ev <- grep("kNN", names(models), value = T, ignore.case = T)
for (o in ev){
  set.seed(22)
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```

### Hyperparameters variations
Let's try increase the range of k values.
```{r nn_tune}
set.seed(22)
tr_control <- trainControl(method = "repeatedcv",
                           repeats = 10,
                           number = 10, # increase number of repetitions
                          classProbs = T) # take into account class probabilities

knn <- train(Y ~ ., data = train,
             method = "knn",
             trControl = tr_control,
             tuneLength = 30) # try 30 values for k
knn
models[["kNN2"]] <- knn

# plot
plot(knn)
```

#### Performance reevaluation
```{r knn2_eval}
ev <- grep("kNN2", names(models), value = T, ignore.case = T)
for (o in ev){
  set.seed(22)
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```


```{r summary_knn}
sum_AcK("kNN", df_cm)
```



## Naive Bayes
we use `Caret` package to apply Naive Bayes algorithm, this depends on `naivebayes` package. 
Let's explore activation or not of *laplace*.

### Data transformation
```{r}
train <- train0
test <- test0
train[,-1] <- scale(train[,-1])
test[,-1] <- scale(test[,-1])
```

### Model training
```{r nb}
tg <- expand.grid(usekernel = T,
                  # laplace value of 0 means no activation
                         laplace = c(0, 0.5, 1 ,1.5), 
                         adjust = 1)

set.seed(22)
tr_control <- trainControl(method="repeatedcv", # method crossvalidation
                           repeats=3) # 10-fold crossvalidation

nb <- train(Y ~ ., data = train,
             method = "naive_bayes",
             trControl = tr_control,
             tuneGrid = tg)
nb
models[["NB"]] <- nb

# plot
plot(nb)
```



### Prediction and evaluation of the model
```{r nb_eval}
ev <- grep("NB", names(models), value = T, ignore.case = T)
for (o in ev){
  set.seed(22)
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```

### Hyperparameters variations
In order to improve the mode we try to increase the values of `laplace`, and using or not `kernel`.

```{r nb2}
tg <- expand.grid(usekernel = c(T,F),
                         laplace = 0:5, 
                         adjust = 0:5)

set.seed(22)
tr_control <- trainControl(method="repeatedcv", # method crossvalidation
                           repeats=3) # 10-fold crossvalidation

nb <- train(Y ~ ., data = train,
             method = "naive_bayes",
             trControl = tr_control,
             tuneGrid = tg)
nb
models[["NB2"]] <- nb

# plot
plot(nb)
```

#### Performance reevaluation
```{r nb2_eval}
ev <- grep("NB2", names(models), value = T, ignore.case = T)
for (o in ev){
  set.seed(22)
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}

```


```{r summary_nb}
sum_AcK("NB", df_cm)
```



## Artificial Neural Network
We will explore dense layers with two hidden layers:  1) 20
y 10 nodes, 2) 50 y 10 nodes.

### Data transformation
On this occasion, we do need to transform the data. We transform the class values into *one hot* codification, needed for neural networks.


```{r}
train <- train0
test <- test0
train[,-1] <- scale(train[,-1])
test[,-1] <- scale(test[,-1])

# to_categorical enables the one hot transformation
train$Y <- train$Y %>% as.numeric()  %>% to_categorical()
test$Y <- test$Y %>% as.numeric() %>% to_categorical()

train <- as.matrix(train)
test <- as.matrix(test)
```

### Model design
We design the model using the `keras` package.

```{r ANN, fig.width=10, fig.height=10}
nodes <- c(20,50)

for (o in nodes){
  tensorflow::set_random_seed(22)
  ann <- keras_model_sequential() 
  ann %>%
    layer_dense(units = o, activation = "relu",
                input_shape = c(200)
                ) %>%
    layer_dropout(rate = 0.5) %>%
    #input shape we indicate the number of genes to analyze
    layer_dense(units = 10, activation = "relu") %>%
    layer_dropout(rate = 0.4) %>%
    layer_flatten() %>% 
    layer_dense(units = 4, 
                # in the final layer we must assign the number of classes
                activation = "softmax" )

  print(ann)
  
  ann %>% compile(loss = "categorical_crossentropy",
                  optimizer = optimizer_rmsprop(),
                  metrics = c("accuracy")
                  )
  models[[paste0("ANN_",o)]] <- ann
}
```

### Training, prediction and evaluation of the model
```{r ann_training_eval}

ev <- grep("ANN", names(models), value = T, ignore.case = T)
for (o in ev){
  tensorflow::set_random_seed(22)
  history <- models[[o]] %>%
              fit(train[,-c(1:4)], train[,1:4], 
                epochs = 30,
                validation_split = 0.2
            )
  print(plot(history)+
          ggtitle(paste0(o, " model"))+
          ggprism::theme_prism() +
          theme(plot.title = element_text(hjust = 0.5))
        )
  
  pred <- models[[o]] %>% predict(test[,-c(1:4)])  %>%
    # revert one hot encoding to see classes
            k_argmax() %>% as.vector() %>% 
            factor(levels=c(1,2,3),
                    labels=c("Basal", "Her2", "LumA"))
  
  # confusion matrix with Caret
  cm <- confusionMatrix(pred, test0[,1])
  cat("##########  ", o, "  ############\n")
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```



### Hyperparameters variations
We add more layers to the previous alrogithm, increase number of nodes and reduce the dropout rate.
```{r ANN2, fig.width=10, fig.height=10}
tensorflow::set_random_seed(22)

ann <- keras_model_sequential() 
ann %>%
  layer_dense(units = 200, activation = "relu",
              input_shape = c(200)
              ) %>%
  layer_dense(units = 50, activation = "relu") %>%
  layer_dropout(rate = 0.2) %>%
  layer_dense(units = 15, activation = "relu") %>%
  layer_dropout(rate = 0.2) %>%
  layer_flatten() %>% 
  layer_dense(units = 4, activation = "softmax" )

print(ann)

ann %>% compile(loss = "categorical_crossentropy",
                optimizer = optimizer_rmsprop(),
                metrics = c("accuracy")
              )
models[[paste0("ANN2")]] <- ann

```


#### Training, prediction and reevaluation of the algorithm
```{r ann2_training_eval}

ev <- grep("ANN2", names(models), value = T, ignore.case = T)
for (o in ev){
  tensorflow::set_random_seed(22)
  history <- models[[o]] %>%
              fit(train[,-c(1:4)], train[,1:4], 
                epochs = 30,
                validation_split = 0.2
            )
  print(plot(history)+
          ggtitle(paste0(o, " model"))+
          ggprism::theme_prism() +
          theme(plot.title = element_text(hjust = 0.5))
        )
  
  pred <- models[[o]] %>% predict(test[,-c(1:4)])  %>%
            k_argmax() %>% as.vector() %>% # revert one hot encoding
            factor(levels=c(1,2,3),
                    labels=c("Basal", "Her2", "LumA"))
    # confusion matrix with caret
  cm <- confusionMatrix(pred, test0[,1])
  cat("##########  ", o, "  ############\n")
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```

```{r summary_ann}
sum_AcK("ANN", df_cm)
```



## Support Vector Machine
Here we will explore the function of lineal and gaussian kernels.

### Data transformation

```{r}
train <- train0
test <- test0
train[,-1] <- scale(train[,-1])
test[,-1] <- scale(test[,-1])
```

### Model training
```{r svmlinear}
set.seed(22)

tr_control <- trainControl(method="repeatedcv", # method crossvalidation
                           repeats=3) # 3-fold crossvalidation
svmlin <- train(Y ~ ., 
                data=train,
                method="svmLinear",
                trControl=tr_control
                ) 
svmlin
models[["SVM_lineal"]] <- svmlin
```

```{r svmRadial, fig.width=10, fig.height=10}
set.seed(22)

#Train model
svmrad <- train(Y ~ ., 
                data=train,
                method="svmRadial",
                trControl=tr_control, # same as previous lineal model
                ) 
svmrad
plot(svmrad)
models[["SVM_RBF"]] <- svmrad
```

### Prediction and evaluation of the model

```{r Eval_SVM}
ev <- grep("SVM", names(models), value = T, ignore.case = T)
for (o in ev){
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```


### Hyperparameters variations

```{r svmlinear2, fig.width=10, fig.height=10}
set.seed(22)

# We try tuning the C values close the ones obtained previously
C <- models$SVM_lineal$bestTune %>% as.integer()
cs <- expand.grid(C = seq(C*0.8,C*1.2,C/20)) 
                                 

svmlin <- train(Y ~ ., 
                data=train,
                method="svmLinear",
                trControl=tr_control,
                tuneGrid = cs 
                ) 
svmlin
plot(svmlin)
models[["SVM2_lineal"]] <- svmlin
```


```{r svmRadial2, fig.width=10, fig.height=10}
set.seed(22)
# We try tuning the C and sigma values close the ones obtained previously
C <- models$SVM_RBF$bestTune[,"C"]
sigma <- models$SVM_RBF$bestTune[,"sigma"]
cs <- expand.grid(C = seq(C*0.7,C*1.3,C/20),
                  sigma = seq(sigma*0.7,sigma*1.3,sigma/20)) 

#Train model
svmrad <- train(Y ~ ., 
                data=train,
                method="svmRadial",
                trControl=tr_control, # same as previous lineal model
                tuneGrid = cs # Different values for C and sigma
                ) 
svmrad
plot(svmrad)
models[["SVM2_RBF"]] <- svmrad
```


#### Performance reevaluation
```{r Eval_SVM2}
ev <- grep("SVM2", names(models), value = T, ignore.case = T)
for (o in ev){
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```

```{r summary_SVM}
sum_AcK("SVM", df_cm)
```


## Trees
We will use the decision tree algorithm *C5.0* using the `C50` package.
We will explore activation or not of the boosting.

### Data transformation
```{r}
train <- train0
test <- test0
train[,-1] <- scale(train[,-1])
test[,-1] <- scale(test[,-1])
```

### Model training
```{r tree, fig.width=20, fig.height=13}
for (o in c(1,10)){ 
  set.seed(22)
  # Loppp with noboost (1 trial) 
  # or with 10 trials (boosting)
  if(o == 1){ex <- "No boosting"
  } else {ex <- paste0(o, " trials of boosting")}
  
  cat("#################", ex, "######################")
  tree <- C5.0(train[,-1], train[,1], trials = o)
  print(tree)
  models[[paste0("Tree.B_",o)]] <- tree
  
  # plot
  print(summary(tree))
  plot(tree, main=ex)
}
```


### Prediction and evaluation of the model
```{r Eval_tree}
ev <- grep("Tree", names(models), value = T, ignore.case = T)
for (o in ev){
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```

### Hyperparameters variations
For the analysis of this data with *decision trees* we will introduce costs to some errors.
As we are dealing with types of tumors here, let's give a high cost to predictions of actual tumors to basal. We will also penalize as prediction an actual basal condition.

```{r matrix_cost}
matrix_dim <- list(levels(rna$Y), levels(rna$Y))
names(matrix_dim) <- c("predicted", "actual")
cost <- matrix(c(0,5,5,
                 10,0,2,
                 10,2,0),
               nrow=3, dimnames = matrix_dim)
cost
```

```{r tree2, fig.width=20, fig.height=13}
for (o in c(1,10)){
  set.seed(22)
  # Loppp with noboost (1 trial) 
  # or with 10 trials (boosting)
  if(o == 1){ex <- "No boosting"
  } else {ex <- paste0(o, " trials of boosting")}
  
  cat("#################", ex, "######################")
  tree <- C5.0(train[,-1], train[,1],
               trials = o, costs = cost)
  print(tree)
  models[[paste0("Tree2.B_",o)]] <- tree
  
  # plot
  print(summary(tree))
  plot(tree, main=ex)
}
```

#### Performance reevaluation
```{r Eval_tree2}
ev <- grep("Tree2", names(models), value = T, ignore.case = T)
for (o in ev){
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```

```{r summary_tree}
sum_AcK("Tree", df_cm)
```


## Random Forest
We will explore different number of tress: n = 50, 100.

### Data transformation

```{r}
train <- train0
test <- test0
train[,-1] <- scale(train[,-1])
test[,-1] <- scale(test[,-1])
```

### Model training
We will use the function `rf`(Random forest) from `Caret`. 

```{r rf}
trees <- c(50,100) # values to try

set.seed(22)
tr_control <- trainControl(method="repeatedcv", # method crossvalidation
                           repeats=3) # 3-fold crossvalidation
for (nt in trees){
  rf <- train(Y ~ ., data = train,
               method = "rf",
               trControl = tr_control,
               ntree = nt)
  cat("##########  ", nt, " arboles ############\n")
  print(rf)
  cat("\n\n")
  models[[paste0("RF_",nt)]] <- rf
  
  # plot
  print(plot(rf))
}
```

### Prediction and evaluation of the model
```{r RF_eval}
ev <- grep("RF", names(models), value = T, ignore.case = T)
for (o in ev){
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```

### Hyperparameters variations
We will use different values for *mtry* close to the best value obtained previously.

```{r rf_2}
trees <- c(50,100) 
mt <- c("50" = models$RF_50$bestTune %>% as.numeric(),
        "100" = models$RF_100$bestTune %>% as.numeric()) 

for (nt in trees){
  set.seed(22)
  m <- mt[[as.character(nt)]]
  cat("##########  ", nt, " arboles ############\n")
  rf <- train(Y ~ ., data = train,
               method = "rf",
               trControl = tr_control, # same as earlier
               tuneGrid = data.frame(mtry = seq(m*0.7,m*1.3,m/20)),
               ntree = nt)
  print(rf)
  cat("\n\n")
  models[[paste0("RF2_",nt)]] <- rf
  
  # plot
  print(plot(rf))
}
``` 

#### Performance reevaluation
```{r RF2_eval}
ev <- grep("RF2", names(models), value = T, ignore.case = T)
for (o in ev){
  cat("##########  ", o, "  ############\n")
  pred <- predict(models[[o]], test)
  cm <- confusionMatrix(pred, test$Y)
  print(cm)
  df_cm <- rbind(df_cm, store_cm(cm, o))
  cat("\n\n")
}
```

```{r summary_RF}
sum_AcK("RF", df_cm)
```


# Model summary
We plot and compare the performance of the different models trained using bar plots.

## Accuracy y kappa for each model.
```{r plot models, fig.width=18, fig.height=9}
# short values to see better differences of high accuracy and kappa
mi <- min(df_cm[df_cm$class == "Overall", "value"])*0.95

#order models
df_cm$Model <- factor(df_cm$Model,
                      levels=names(models))

df_cm %>% filter(class == "Overall") %>%
  ggplot(aes(Model, value, fill=metric)) +
  geom_col()+
  geom_text(aes(label=round(value, 2),
                fontface="bold"))+
  coord_cartesian(ylim=c(mi,1))+
  facet_wrap(.~ metric) +
  ggprism::theme_prism() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Define metrics to visualize for each class
metri <- c("Sensitivity", "Specificity",         
          "Pos Pred Value", "Neg Pred Value",
          "Precision", "Detection Rate")
```

Next we examine the `r paste(metri, collapse=", ")` for each type of class.

```{r plot models2, fig.width=22, fig.height=16}
df_cm %>% filter(class != "Overall" &
                metric %in% metri) %>%
  separate(class, into=c("out","class"),
           sep = " ", remove = T) %>% 
  ggplot(aes(class, value, fill=Model)) +
  geom_bar(stat="identity", position = "dodge")+
  facet_wrap( .~ metric, ncol = 2) +
  ggprism::theme_prism()

```

Finally, we export a .xlsx file with the summary of the models.

```{r export_df, include=F}
df_cm %>%
  pivot_wider(names_from = Model, values_from = value) %>%
  writexl::write_xlsx("Resumen_modelos.xlsx")
```
